--- 
title: "Deaths related to the Volkswagen defeat devices"
output:
  html_document:
    toc: true
    number_sections: true
    theme: cosmo
    toc_depth: 2
---




```{r lib, warning =  F, message = F, echo = F}
# File to test mcmc_corr.R
library(mvtnorm)
library(truncnorm)
library(MESS)
library(knitr)
library(dplyr)
library(tidyr)
opts_chunk$set(message = F, fig.align = "center", warning = F, echo = F)
```

# Introduction

This report attempts to reproduce analyses conducted by the New York Times, Vox, Mother Jones, and the Associated Press to determine the number of deaths associated with the Volkswagen defeat devices.

Here are important statistics used in some of the analyses:

- Approximate number of affected Volkswagen vehicles in the US: [482000](http://yosemite.epa.gov/opa/admpress.nsf/6424ac1caa800aab85257359003f5337/dfc8e33b5ab162b985257ec40057813b!OpenDocument) (source:  US EPA)
- Approximate number of affected Volkswagen vehicles in California: [50000](http://www.arb.ca.gov/newsrel/in_use_compliance_letter.htm/) (source: California Air Resources Board)
- Average number of miles driven per year in the US: [11244](http://www.afdc.energy.gov/data/10309/) (source: US Department of Energy) 
- EPA Tier 2 Bin 5 standard at useful life: [0.07 grams / mile](http://www3.epa.gov/otaq/standards/light-duty/tier2stds.htm) (source: US EPA)
- EPA Tier 2 Bin 5 standard at 50,000 miles: [0.05 grams / mile](http://www3.epa.gov/otaq/standards/light-duty/tier2stds.htm) (source: US EPA)
- Amount the affected Volkswagen vehicles exceeded the US EPA standards of 0.07 for NOx emissions: [15-35 times](http://www.theicct.org/news/epas-notice-violation-clean-air-act-volkswagen-press-statement) (source: The International Council on Clean Transportation) 


```{r, echo = F}
#rounding 
r1 <- 0

#initialize estimates
dat <- data.frame(matrix(nrow = 6, ncol = 7))
colnames(dat) <- c("estimate", "miles", "exceed", "exnox", "deathston", "poll", "deaths")
dat$estimate <- c("New York Times: MSK-JS", "New York Times: Selin", "Vox", "Mother Jones", "Associated Press", "JRK")
```


# New York Times

[Article link](http://www.nytimes.com/2015/09/29/upshot/how-many-deaths-did-volkswagens-deception-cause-in-us.html)

## Estimate 1: Margot Sanger-Katz and John Schwartz

```{r}
# This is the sum of affected Volkswagen vehicles sold in each each from 2009 until 2015
numcars <- 461723 - 13

# It is not clear how this estimate was obtained
milestot <- 23463225312
dat$miles[1 : 2] <- milestot / 10^9

# The US EPA limit at 50,000 miles
limit <- 0.05
# 40 times the limit
exlimit <- 39

# Grams per ton
gramton <- 907185

# These computations do not match those from NYT
# Tons emittted
tons <- milestot * limit * exlimit / gramton
# Tons emitted by season
tonssea <- tons * 5/12

# total tons in US since late 2008
tons <- 46000


# Unclear how this estimate was obtained, from  Dechênes et al. working paper
deaths1 <- 106
dat$deaths[1] <- deaths1


dat$exceed[1] <- 40
dat$exnox[1] <- tons
dat$deathston[1] <- ""
dat$poll[1] <- "All from power plants"
```

The authors estimated the number of affected vehicles in the US to be approximately 460,000 by excluding cars sold in July and 2016 models. They estimated the number of miles driven by affected cars to be 23,463,225,312 and the excess NOx to be 46,000 tons.  The number of deaths, 106, was obtained using estimates from the Dechenes et al. working paper cited in the article, which estimated the benefits of reducing NOx from power plant emissions. 

## Estimate 2: Noelle Eckley Selin

```{r}
# total tons in US since late 2008 (see above from Sanger-Katz/Schwartz estimate)
tons <- 46000

# Deaths per ton of NOx from EPA/ Krewski et al. 2009
deathsperton <- 0.00085
deaths <- deathsperton * tons

# Total deaths over 6 years in US
deaths1 <- round(deaths, r1)



dat$exceed[2] <- 40
dat$exnox[2] <- tons
dat$deathston[2] <- deathsperton
dat$deaths[2] <- deaths1
dat$poll[2] <- "PM2.5"
```

This estimate was obtained using the same number of miles and estimated excess tons of NOx as the Sanger-Katz/Schwartz estimate. To obtain the estimated number of deaths, they used the estimated increase in deaths per ton of NOx related to particulate matter (PM) from the EPA technical support document "Estimating the Benefit per Ton of Reducing PM2.5 Precursors from 17 Sectors". 
 

# Vox

[Article link](http://www.vox.com/2015/9/23/9383641/volkswagen-scandal-pollution/)

```{r}
# Number of VW cars sold in US (total)
cars <- 482000
# US EPA Tier 2- Bin 5 NOx emission standard in grams per mile
limit <- 0.07
# VW cars exceeded legal limits by X times
exlimit <- c(15, 35)
# Deaths per ton of NOx from EPA technical support document
deathperton <- c(0.00085, 0.0019)
# Miles driven per year (average)
milesyear <- 11244
# Grams per ton
gramton <- 907185

# Total miles driven
miles <- milesyear * cars


# Range of excess NOx emissions
exnox <- (limit * (exlimit - 1)) * cars * milesyear / gramton

# Range of excess deaths per year in US
deaths <- exnox * deathperton

# Range of excess deaths in US over 6 years
deaths1 <- round(deaths * 6, r1)



dat$miles[3] <- miles * 6 / 10^9
dat$exceed[3] <- "15-35"
dat$exnox[3] <- paste0("(", paste(round(exnox * 6, r1), collapse = "-"), ")")
dat$deathston[3] <- "0.00085-0.0019"
dat$deaths[3] <- paste0("(", paste(deaths1, collapse = "-"), ")")
dat$poll[3] <- "PM2.5"
```

The Vox analysis estimated between 5,800 and 14,200 excess tons of NOx were emitted per year, by assuming that each vehicle drove on average 11,244 miles per year and the NOx emissions were 15-35 times the EPA standard of 0.07 grams per mile.    

To estimate the number of deaths, they used the estimated increase in deaths per ton of NOx related to particulate matter (PM) from the EPA technical support document "Estimating the Benefit per Ton of Reducing PM2.5 Precursors from 17 Sectors".  This document provided two estimates: 0.00085 deaths per ton of NOx and 0.0019 deaths per ton of NOx.  The authors estimated 5-27 deaths per year in the US were related to the affected Volkswagen vehicles.  If we assume that the emissions were over 6 years (to match estimates from Mother Jones), the estimated deaths range from 30-162 in the US. 
 


# Mother Jones

[Article link](http://www.motherjones.com/kevin-drum/2015/09/spreadsheet-day-how-many-people-did-vw-kill/)

```{r}
# Number of VW cars sold in CA
carsCA <- 50000

# Excess tons of NOx over 6 years
milesyear <- 12000

# Approximate increasing number of cars sold to equal 50000
cars <- c(3000, 5000, 7000, 10000, 12000, 13000)
years <- seq(2009, 2014)
miles <- milesyear * (2015 - years)

# Total miles driven in CA over 6 years
milestot <- sum(cars * miles)



# Excess NOx per mile 
exnoxmil <- 30 * 0.07

# Grams per ton
gramton <- 907185

# Excess total NOx in tons
exnox <- exnoxmil * milestot / gramton

# Deaths per ton (range of 0.0019, 0.0095)
# Not clear where .0044 comes from, range from (Mauzerall et al. 2005)
deathsperton <- c(0.0019, 0.0044, 0.0095)

# Estimated excess deaths in California
deaths <- exnox * deathsperton

# Estimated excess deaths in US over 6 years
# Obtained by 50000 approximately 1/10 total cars sold
deaths1 <- round(deaths * 10, r1)


# Significance table
# Total miles driven in US over 6 years
dat$miles[4] <- milestot / 10^9 * 10
dat$exceed[4] <- 30
dat$exnox[4] <- round(exnox * 10, r1)
dat$deathston[4] <- "0.0019-0.0095"
dat$deaths[4] <- paste0("(", deaths1[1], "-", deaths1[3], ")")
dat$poll[4] <- "ozone"
```


The Mother Jones analysis assumed that the number of affected vehicles on the road increased from 2009 through 2015, since Model Years 2009-2015 were impacted.  Assuming an average of 12000 miles were driven per year and their analysis, I estimate a total of 16.6 billion miles driven by affected vehicles in the US.

They assumed that vehicles were 30 times the EPA limit of 0.07.  To estimate the number of deaths, they used 0.0044 deaths per ton of NOx, which is derived from the range of deaths per ton of NOx related to ozone exposure of 0.0019-0.0095 from Mauzerall et al. (2005).  This gives a point estimate of 17 deaths in California, with 170 deaths (range of 73-364 deaths) in the US, by assuming California has 1/10 of the US population.     

# Associated Press

[Article link](http://bigstory.ap.org/article/a6925f0af82e44aaa1a1ed4b55d030f6/ap-analysis-dozens-deaths-likely-vw-pollution-dodge)


```{r}





#12 to 69 deaths or 16-94 deaths.
dat$deaths[5] <- "(16-94)"
``` 

Little information was provided about the assumptions used in the AP analysis. They reported 16-94 deaths over 7 years.   



# Significance article

```{r}

dat$miles <- round(dat$miles, r1)

cars <- 482000
milesyear <- 11244

miles1 <- (cars * milesyear * c(1, 6) / 10^9) %>% round(., r1)
dat$miles[6] <- paste0("(", paste(miles1, collapse = "-"), ")")


dat <- dat[c(1, 2, 3, 4, 5), ] 

colnames(dat) <- c("Source", "Miles (billions)", "Amount NOx exceed limit", "Excess NOx (tons)", "Deaths per ton of NOx", "Ambient pollutant causing death", "Total US Deaths")

kable(dat, row.names = F)


write.csv(dat, file = "vw-estimates.csv", row.names = F)
```


To determine the greatest source of uncertainty, we used baseline values of 17 billion miles driven, affected vehicles were 15 times the EPA limit of 0.07, and 0.00085 deaths per ton of NOx.  Then, the smallest estimated number of deaths given the range of values used in the analyses was 15 deaths. Changing the number of deaths per ton of NOx to the largest value, 0.0095, yields 174 deaths.  Changing the number of miles driven to 33 billion, we estimate 30 deaths.  Changing the amount by which the Volkswagen vehicles exceeded the limit from 15 to 40 gives 43 deaths. Therefore the largest source of uncertainty related to the Volkswagen emissions is the estimated deaths per ton of NOx emitted.    


```{r, eval = T}
# Uncertainty evaluation
miles <- c(6, 17, 33)
exceed <- c(15, 30, 40)
deathton <- c(0.00085, 0.0019, 0.0095)
dat2 <- expand.grid(miles, exceed, deathton)
colnames(dat2) <- c("miles", "exceed", "deathton")

# Compute deaths
dat2 <- mutate(dat2, bl = round(miles * 10^9 *0.07 * (exceed - 1) / 907185 * deathton))

# Plot histogram of deaths
dat2$miles <- factor(dat2$miles)
dat2$exceed <- factor(dat2$exceed)
dat2$deathton <- factor(dat2$deathton)
g1 <- ggplot(dat2, aes(x = bl)) + 
  geom_histogram(alpha = 0.5, color = "black") +
  theme_bw() + xlab("Estimated US deaths") + 
  ylab("Number of scenarios") +
  scale_fill_brewer(palette = "Blues", name = "Deaths per ton") +
  #theme(text = element_text(size = 14)) 
png("vw-unc-plot.png", height = 350 * 2, width = 480 * 2)
g1
dev.off()

pdf("vw-unc-plot.pdf", height = 3, width = 4)
g1
dev.off()

embedFonts("vw-unc-plot.pdf")


#dat2 <- spread(dat2, deathton, bl)

# Save table
#colnames(dat2) <- c("Miles (billions)", "Amount NOx exceed limit", "Deaths per ton of NOx", "Total US Deaths")

#kable(dat2)
#dat2[c(1, 3, 7, 19), ]
dat2 <- dat2[c(14, 15, 17, 23), ]

write.csv(dat2, file = "est-uncertainty.csv", row.names = F)
```


