--- 
title: "Volkswagen Estimates"
output:
  html_document:
    toc: true
    number_sections: true
    theme: cosmo
    toc_depth: 2
---




```{r lib, warning F, message F, echo = F}
# File to test mcmc_corr.R
library(mvtnorm)
library(truncnorm)
library(MESS)
library(knitr)


opts_chunk$set(message = F, fig.align = "center", warning = F, echo = T)
```

# Overall

- Number of affected VW cars in US estimated by the EPA is [482000](http://yosemite.epa.gov/opa/admpress.nsf/6424ac1caa800aab85257359003f5337/dfc8e33b5ab162b985257ec40057813b!OpenDocument/)
- Number of affected VW cars in California estimated by the California Air Resources Board is [50000](http://www.arb.ca.gov/newsrel/in_use_compliance_letter.htm/)
- Average number of miles driven per year estimated by the US Department of Energy to be [11244](http://www.afdc.energy.gov/data/10309/) 
- EPA Tier 2 Bin 5 standard [0.07 grams / mile](http://www3.epa.gov/otaq/standards/light-duty/tier2stds.htm/)
- ICCT estimated VW cars had NOx emissions [15-35 times](http://www.theicct.org/news/epas-notice-violation-clean-air-act-volkswagen-press-statement) the legal limit (0.07) 

# New York Times

[Article link](http://www.nytimes.com/2015/09/29/upshot/how-many-deaths-did-volkswagens-deception-cause-in-us.html/)

## Estimate 1: Margot Sanger-Katz and John Schwartz

```{r}
# total tons in US since late 2008
tons <- 46000

350 / (46000 / 6)
46000 / (.38 * 1000 * 2539)

# unclear, from  Dechênes working paper
redmort <- .04

deaths <- 106
```

## Estimate 2: Noelle Eckley Selin

```{r}
# total tons in US since late 2008
tons <- 46000

# Deaths per ton of NOx from EPA/ Krewski et al. 2009
deathsperton <- 0.00085

deaths <- deathsperton * tons

# Total deaths over 6 years in US
round(deaths, 2)
```


 

# Vox

[Article link](http://www.vox.com/2015/9/23/9383641/volkswagen-scandal-pollution/)

```{r}
# Number of VW cars sold in US (total)
cars <- 482000
# Tier 2- Bin 5 NOx emission standard in grams per mile
# (From EPA limits)
limit <- 0.07
# VW cars exceeded legal limits by X times
# From 
exlimit <- c(15, 35)
# Deaths per ton of NOx
deathperton <- c(0.00085, 0.0019)
# Miles driven per year (average)
milesyear <- 11244
# Grams per ton
gramton <- 907185

# Range of excess NOx emissions
exnox <- (limit * (exlimit - 1)) * cars * milesyear / gramton

# Range of excess deaths per year in US
deaths <- exnox * deathperton

# Range of excess deaths in US over 6 years
round(deaths * 6, 2)
```




# Mother Jones

[Article link](http://www.motherjones.com/kevin-drum/2015/09/spreadsheet-day-how-many-people-did-vw-kill/)

```{r}
# Number of VW cars sold in CA
# From CARB compliance letter
carsCA <- 50000

# Excess tons of NOx over 6 years
# miles driven per year (approximate with 11244 from Vox?)
milesyear <- 12000

# Approximate increasing number of cars sold to equal 50000
cars <- c(3000, 5000, 7000, 10000, 12000, 13000)
years <- seq(2009, 2014)
miles <- milesyear * (2015 - years)

# Total miles driven over 6 years
milestot <- sum(cars * miles)

# Excess NOx per mile 
exnoxmil <- 30 * 0.07

# Grams per ton
gramton <- 907185

# Excexx total NOx in tons
exnox <- exnoxmil * milestot / gramton

# Deaths per ton (range of 0.0019, 0.0095)
# Not clear where .0044 comes from, range from (Mauzerall et al. 2005)
deathsperton <- c(0.0019, 0.0044, 0.0095)

# Estimated excess deaths in California
deaths <- exnox * deathsperton

# Estimated excess deaths in US over 6 years
# Obtained by 50000 approximately 1/10 total cars sold
round(deaths * 10, 2)
```
